Type theory with single substitutions

### 2025-02-21 cpp/p.tex
spellchec

However, even at the GAT level of abstraction, there are some choices
to be made about the instantiation operation. It is usually convenient
to use parallel substitutions, which means that the instantiation
operation replaces all variables in a context by terms in another
context \emph{at the same time}, that is, $t[x_0\mapsto
  u_0,\dots,x_n\mapsto u_n]$ where $x_0,\dots,x_n$ are the variables
in the context of $t$. $(x_0\mapsto u_0,\dots,x_n\mapsto u_n)$ is
called a parallel substitution, where the terms $u_0,\dots,u_n$ are
all in the same context. Parallel substitutions are very natural in
the algebraic setting because they form the morphisms in a category
where objects are contexts (lists of types), then types/terms are
preheaves, context extension is a representability condition, so the
equations of the algebraic theory are forced by the categorical
structure. Some examples of algebraic definitions of type theory using
parallel substitutions ranging from the more syntactic to the more
semantic: Ehrhard's notion \cite{ehrhard,coquandEhrhard}, categories
with families (CwF \cite{DBLP:conf/types/Dybjer95,Castellan2021}),
Awodey's natural models \cite{DBLP:journals/mscs/Awodey18}, contextual
categories \cite{DBLP:journals/apal/Cartmell86}, C-systems
\cite{DBLP:journals/lmcs/AhrensLV18}, locally cartesian closed
categories \cite{DBLP:journals/mscs/ClairambaultD14}, path categories
\cite{10.1145/3204492}.

In traditional extrinsic presentations
(e.g.\ \cite{alti:phd93,DBLP:books/daglib/0005958}) instantiation
replaces free occurrences of a single variable $x$ with a term $u$ as
in the notation $t[x\mapsto u]$. A substitution consists of a
variable name (a pointer into a context) and a term. Substitutions
cannot be composed, thus they do not form a category. To bridge the
gap between the intrinsic parallel substitution world and the
extrinsic single substitution world, Voevodsky introduced B-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023} which are an algebraic
desription of a single substitution calculus. B-systems involve
complex rules describing telescopes, weakenings and substitutions
under telescopes, with several equations.

@@

Our metatheory is observational type theory
\cite{DBLP:phd/hal/Pujet22} with quotient inductive-inductive types
(QIITs). On paper we usually omit writing coercions, so our notation
is close to extensional type theory. Sections \ref{sec:tt} and
\ref{sec:admissible} of this paper are formalised in the proof
assistant Agda. The formalisation is availble at TODO \href{https://szumixie.github.io/single-subst/TT.index.html}{\Agda}. We use a strict
Prop-valued \cite{DBLP:journals/pacmpl/GilbertCST19} equality type
with postulated coercion rule (transport rule) and postulated QIITs
with computation rules added using rewrite rules
\cite{DBLP:journals/pacmpl/CockxTW21}. Thus, we work in a setting of
uniqueness of identity proofs (UIP), and our metatheory is
incompatible with homotopy type theory \cite{HoTTbook}. Our notation
on paper is close to Agda's: we write $\Pi$ types as $(x:A)\ra B$,
$\Sigma$ types as $(x:A)\times B$, we use implicit arguments and
overloading extensively, for example instantiation $\blank[\blank]$ is
overloaded for types and terms. For isomorphisms, we use the notation
$f:X\cong Y:g$ meaning $f:X\ra Y$ and $g :Y\ra X$ together with a
$\beta$ equality $f\,(g\,y) = y$ for all $y$, and an $\eta$ equality
saying $g\,(f\,x) = x$ for all $x$ (we think of $f$ as a destructor
and $g$ as a constructor).

@@

First of all, we need a sort of terms which have to be indexed by
types because we only want well-typed terms. Both types and terms can
include variables, and to keep track of the currently available
variables, we also index them by contexts: a context is a list of
types, the length is the number of available variables, and we add new
variables at the end (that is, they are snoc-lists rather than
cons-lists). Types are also indexed by their universe level, this is a
technical requirement for avoiding Russell's paradox. The index $i$ is
an implicit argument of $\Tm$.
\[
\Con : \Set \hspace{3em} 
\Ty : \Con\ra\N\ra\Set \hspace{3em} 
\Tm : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Set
\]
Just as lists have two constructors, there are two ways to form
contexts: the empty context $\diamond$ and context extension $\ext$
which is like the snoc operation for lists. Context extension takes a
type which can have variables in the context preceeding the type.
\[
\diamond : \Con \hspace{3em}
\blank\ext\blank : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Con
\]
We will refer to variables by their distance from the end of the
context. We would like to describe the operation providing the last
variable in a context, the zero De Bruijn index \cite{debruijn}, we
denote it by $\q : \Tm\,(\Gamma\ext A)\,A'$. It is a term in an
extended context, and it should have type $A$, but the issue is that
$A : \Ty\,\Gamma$, and $A' : \Ty\,(\Gamma\ext A)$. We need a way to
weaken the type $A$ to obtain such an $A'$. For this, we will
introduce an operation $\blank[\p] : \Ty\,\Gamma\,i\ra\Ty\,(\Gamma\ext
A)\,i$ and define $A' := A[\p]$. Instead of introducing just
$\blank[\p]$, we generalise a bit and add a new sort $\Sub$ which for
now only has the single element $\p$ and the operation
$\blank[\blank]$ is called \emph{instantiation}. We instantiate the
type with the $\Sub\,\Delta\,\Gamma$ weakening.  Elements of $\Sub$
will later be called substitutions, hence the name, but for now, we
only have single end-of context weakenings in $\Sub$. The context
$\Delta$ is called the domain, $\Gamma$ the codomain of a
$\Sub\,\Delta\,\Gamma$.
\begin{alignat*}{10}
  & \Sub && : \Con\ra\Con\ra\Set && \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
  & \blank[\blank] && : \Ty\,\Gamma\,i\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta\,i \hspace{3em} && \q && : \Tm\,(\Gamma\ext A)\,(A[\p])
\end{alignat*}
The operations $\p$ and $\q$ take three implicit parameters, $\Gamma$,
$i$ and $A$, while $\blank[\blank]$ takes $\Gamma$, $i$ and $\Delta$
implicitly. Still, we only have the last variable $\q$ in the context,
we don't have e.g.\ the last but one variable in $\Tm\,(\Gamma\ext
A\ext B)\,(A[\p][\p])$ where we had to weaken $A$ twice to make it fit
with its context. To obtain more variables, we also allow weakening of
terms, more precisely, we introduce an instantiation operation for
terms with an overloaded notation.
\begin{alignat*}{10}
& \blank[\blank] && : \Tm\,\Gamma\,A\ra(\gamma:\Sub\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\gamma])
\end{alignat*}
Note that this is a dependent function as the type of the resulting
term has to be weakened the same way as the term itself. Now we can
define all variables counting from the end of the context (De Bruijn
indices): $0 := \q$, $1 := \q[\p]$, $2 := \q[\p][\p]$,
$3 := \q[\p][\p][\p]$, and so on.

@@

We add the rules for universes. A universe is a type containing codes
for types, this is witnessed by $\El$ (elements) and $\cd$ (code),
which make an isomorphism between terms of type $\U\,i$ and types of
level $i$. All three operations come with substitution rules.
% ($\El[]$
% and $\cd[]$ are interderivable, so it would be enough to assume one of
% the two, but we add both to completely align with the algorithm
% \cite{DBLP:conf/fscd/KaposiX24} generating the rules from the SOGAT
% (\ref{eq:tt})).
\begin{alignat*}{10}
  & \U && : (i:\N)\ra\Ty\,\Gamma\,(1+i)\hspace{3em} && \U[] && : (\U\,i)[\gamma] = \U\,i                           && \U\beta && : \El\,(\cd\,A) = A \\        
  & \El && : \Tm\,\Gamma\,(\U\,i) \ra \Ty\,\Gamma\,i && \El[] && : (\El\,\hat{A})[\gamma] = \El\,(\hat{A}[\gamma]) \hspace{3em} && \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A} \\
  & \cd && : \Ty\,\Gamma\,i\ra\Tm\,\Gamma\,(\U\,i) && \cd[] && : (\cd\,A)[\gamma] = \cd\,(A[\gamma])
\end{alignat*}
Finally, we add the rules for moving types one level up. Because $\Pi$
needs two types at the same level, without lifting, we cannot even
define the polymorphic identity function.
\begin{alignat*}{10}
  & \Lift && : \Ty\,\Gamma\,i\ra\Ty\,\Gamma\,(1+i)\hspace{3em} && (\Lift\,A)[\gamma] = \Lift\,(A[\gamma])\hspace{3em} && \Lift\beta && : \un\,(\mk\,a) = a \\
  & \mk && : \Tm\,\Gamma\,A \ra \Tm\,\Gamma\,(\Lift\,A) && (\mk\,a)[\gamma] = \mk\,(a[\gamma])              && \Lift\eta && : \mk\,(\un\,a) = a \\
  & \un && : \Tm\,\Gamma\,(\Lift\,A) \ra \Tm\,\Gamma\,A && (\un\,a)[\gamma] = \un\,(a[\gamma])
\end{alignat*}
Note that we don't have definitional commutation of type formers and
$\Lift$ such as $\Lift\,(A\Ra B) = (\Lift\,A\Ra\Lift\,B)$. It however
holds as a definitional isomorphism (see Subsection
\ref{sec:examples}). This adds extra administration when using our
theory, but makes stating our theory simpler, and includes more
models. The situation is similar to the correspondance between
$\Ty\,\Gamma\,i$ and $\Tm\,\Gamma\,(\U\,i)$ which could have been
stated as an equality (making universes {\`a} la Russell).

@@

The operations for $\U$ and $\Lift$ can be written more concisely as
the following isomorphisms:
\[
\El:\Tm\,\Gamma\,(\U\,i) \cong \Ty\,\Gamma\,i:\cd \hspace{5em} \un:\Tm\,\Gamma\,(\Lift\,A)\cong\Tm\,\Gamma\,A:\mk
\]
The $\U\beta$, $\U\eta$ and $\Lift\beta$, $\Lift\eta$ rules express
that the roundtrips are identities. These isomorphisms are also
\emph{natural}, or compatible with substitutions: these are expressed
by the equations $\El[]$, $\un[]$, $\cd[]$, $\mk[]$.

$\Sigma$ types were introduced by a similar isomorphism, and we give
names to the roundtrips equations according to our convention introduced in
Section \ref{sec:metatheory}: $\Sigma\beta_1 : \fst\,(a,b) = a$,
$\Sigma\beta_2 : \snd\,(a,b) = b$ and $\Sigma\eta : w =
(\fst\,w,\snd\,w)$. Naturality for $\Sigma$ was only stated in one
direction (${,}[]$), naturalities in the other direction are
derivable, for $\fst$ this is
$
(\fst\,w)[\gamma]  \overset{\Sigma\beta_1}{=}                          
\fst\,\big((\fst\,w)[\gamma],(\snd\,w)[\gamma]\big) \overset{{,}[]}{=} 
\fst\,\big((\fst\,w,\snd\,w)[\gamma]\big)  \overset{\Sigma\eta}=        
\fst\,(w[\gamma]).
$
For $\Pi$ types, we construct a similar isomorphism
$
(\lambda t.t[\p]\cdot\q) : \Tm\,\Gamma\,(\Pi\,A\,B) \cong \Tm\,(\Gamma\ext A)\,B : \lam.
$
The first roundtrip equality is proven as
$\beta : (\lam\,t)[\p]\cdot\q \overset{\lam[]}{=} \lam\,(t[\p^+])\cdot\q \overset{\Pi\beta}{=} t[p^+][\langle\q\rangle] \overset{\text{(\ref{eq:lifted4})}}{=} t$.
The last step is the admissible equation (\ref{eq:lifted4}) from
Section \ref{sec:admissible}. As above with weakening into an
arbitrary context, this equation holds in the syntax, bot not in an
arbitrary model.
%and also if $t$ is not a metavariable, but an operation, e.g. if $t = \q$ or $t = \q[\p]\cdot\q$).
The other roundtrip is exactly $\Pi\eta : \lam\,(t[\p]\cdot\q) = t$.

We construct another isomorphism between terms depending on a variable and
terms depending on the lifted version of the same variable:
\begin{equation}\label{eq:liftvar}
(\lambda t.t[\p^+][\langle\un\,\q\rangle]) : \Tm\,(\Gamma\ext A)\,B \cong \Tm\,(\Gamma\ext\Lift\,A)\,(B[\p^+][\langle\q\rangle]) : (\lambda t.t[\p^+][\langle\mk\,\q\rangle])
\end{equation}
To derive the roundtrips, we again need admissible equations. We prove
the first roundtrip equality in Appendix \ref{app:liftvar}, the other
is analogous.
Once we have all the above isomorphisms, we can compose them to obtain commutation of $\Lift$ and $\Pi$:
\begin{alignat*}{10}
  & \Tm\,\Gamma\,\big(\Lift\,(\Pi\,A\,B)\big) \overset{\text{$\Lift$}}{\cong}
  \Tm\,\Gamma\,(\Pi\,A\,B) \overset{\text{$\Pi$}}{\cong} 
  \Tm\,(\Gamma\ext A)\,B \overset{\text{$\Lift$}}{\cong} 
  \Tm\,(\Gamma\ext A)\,(\Lift\,B) \overset{\text{lifted var}}{\cong} \\
  & \Tm\,(\Gamma\ext \Lift\,A)\,(\Lift\,B[\p^+][\langle\un\,\q\rangle]) \overset{\text{$\Pi$}}{\cong}
  \Tm\,\Gamma\,\big(\Pi\,(\Lift\,A)\,(\Lift\,B[\p^+][\langle\un\,\q\rangle])\big)
\end{alignat*}
We hope that the above examples demonstrated the need for equations
(\ref{eq:lifted1})--(\ref{eq:lifted4}) which we will prove in the next
section.

@@

We can simulate parallel substitutions using $\Sub^*$ as iterated single
substitutions, however it does not satisfy the equations of parallel
substitutions. Thus we define parallel substitutions as lists of terms. It seems to be
difficult to define the instatiation operation for $\Tms$ directly, so we also define
a map into $\Sub^*$, and reuse the instantiation operation of $\Sub^*$ to
convert it into a sequence of instantiations of single substitutions.
\begin{alignat*}{10}
  & \Tms && : \Con\ra\Con\ra\Con && \llcorner\blank\lrcorner && : \Tms\,\Delta\,\Gamma \to \Sub^*\,\Delta\,\Gamma \\
  & \epsilon && : \Tms\,\Gamma\,\diamond && \llcorner\epsilon\lrcorner && := \id\circ\p\circ\dots\circ\p \\
  & \blank,\blank && : (\gamma:\Tms\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\llcorner\gamma\lrcorner])\ra\Tms\,\Delta\,(\Gamma\ext A)\hspace{1.9em} && \llcorner\gamma,a\lrcorner && := \llcorner\gamma\lrcorner^+\circ\langle a \rangle
\end{alignat*}
The computation of instantiating with $\Tms$ is illustrated below. On the
right-hand side, we first weaken $B$ to the context in which the $a_i$'s live,
then replace the variables of $B$ one by one using the $a_i$'s, which are
well-typed because of the previous weakening.
\[
  B[\llcorner\epsilon,a_1,a_2,\dots,a_{n-1},a_n\lrcorner] = 
  B[\p^{+^n}]\dots[\p^{+^n}][\langle a_1 \rangle^{+^{n-1}}][\langle a_2 \rangle^{+^{n-2}}]\dots[\langle a_{n-1} \rangle^+][\langle a_n \rangle]
\]
All CwF operations and equations can be defined with $\Tms$ by induction, using
Lemma \ref{thm:lifting} to avoid further induction on the syntax.

\begin{problem}
  Contexts, types, and terms in the SSC syntax are isomorphic to the
  corresponding sorts in the CwF syntax. In addition $\Tms$ is isomorphic to CwF
  substitutions.
\end{problem}
\begin{proof}[Construction]\leavevmode
  \begin{description}
    \item[$\Rightarrow$] By recursion on the syntax, SSC operations can be
    trivially interpreted by CwF operations.
    \item[$\Leftarrow$] By recursion on the syntax, using $\Tms$ to interpret
    parallel substitutions.
  \end{description}
  The roundtrips are proven by induction on the syntax.
\end{proof}
This also implies that the initial SSC is the initial CwF.

@@

The other roundtrip, that is, starting with an SSC-model, termifying

@@

isomorphic after the roundtrip. We leave formulating the right notion

>>>>

However, even at the GAT level of abstraction, there are some choices
to be made about the instantiation operation. It is usually convenient
to use parallel substitutions, which means that the instantiation
operation replaces all variables in a context by terms in another
context \emph{at the same time}, that is, $t[x_0\mapsto
  u_0,\dots,x_n\mapsto u_n]$ where $x_0,\dots,x_n$ are the variables
in the context of $t$. $(x_0\mapsto u_0,\dots,x_n\mapsto u_n)$ is
called a parallel substitution, where the terms $u_0,\dots,u_n$ are
all in the same context. Parallel substitutions are very natural in
the algebraic setting because they form the morphisms in a category
where objects are contexts (lists of types), then types/terms are
presheaves, context extension is a representability condition, so the
equations of the algebraic theory are forced by the categorical
structure. Some examples of algebraic definitions of type theory using
parallel substitutions ranging from the more syntactic to the more
semantic: Ehrhard's notion \cite{ehrhard,coquandEhrhard}, categories
with families (CwF \cite{DBLP:conf/types/Dybjer95,Castellan2021}),
Awodey's natural models \cite{DBLP:journals/mscs/Awodey18}, contextual
categories \cite{DBLP:journals/apal/Cartmell86}, C-systems
\cite{DBLP:journals/lmcs/AhrensLV18}, locally cartesian closed
categories \cite{DBLP:journals/mscs/ClairambaultD14}, path categories
\cite{10.1145/3204492}.

In traditional extrinsic presentations
(e.g.\ \cite{alti:phd93,DBLP:books/daglib/0005958}) instantiation
replaces free occurrences of a single variable $x$ with a term $u$ as
in the notation $t[x\mapsto u]$. A substitution consists of a
variable name (a pointer into a context) and a term. Substitutions
cannot be composed, thus they do not form a category. To bridge the
gap between the intrinsic parallel substitution world and the
extrinsic single substitution world, Voevodsky introduced B-systems
\cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023} which are an algebraic
description of a single substitution calculus. B-systems involve
complex rules describing telescopes, weakenings and substitutions
under telescopes, with several equations.

@@

Our metatheory is observational type theory
\cite{DBLP:phd/hal/Pujet22} with quotient inductive-inductive types
(QIITs). On paper we usually omit writing coercions, so our notation
is close to extensional type theory. Sections \ref{sec:tt} and
\ref{sec:admissible} of this paper are formalised in the proof
assistant Agda. The formalisation is available at TODO \href{https://szumixie.github.io/single-subst/TT.index.html}{\Agda}. We use a strict
Prop-valued \cite{DBLP:journals/pacmpl/GilbertCST19} equality type
with postulated coercion rule (transport rule) and postulated QIITs
with computation rules added using rewrite rules
\cite{DBLP:journals/pacmpl/CockxTW21}. Thus, we work in a setting of
uniqueness of identity proofs (UIP), and our metatheory is
incompatible with homotopy type theory \cite{HoTTbook}. Our notation
on paper is close to Agda's: we write $\Pi$ types as $(x:A)\ra B$,
$\Sigma$ types as $(x:A)\times B$, we use implicit arguments and
overloading extensively, for example instantiation $\blank[\blank]$ is
overloaded for types and terms. For isomorphisms, we use the notation
$f:X\cong Y:g$ meaning $f:X\ra Y$ and $g :Y\ra X$ together with a
$\beta$ equality $f\,(g\,y) = y$ for all $y$, and an $\eta$ equality
saying $g\,(f\,x) = x$ for all $x$ (we think of $f$ as a destructor
and $g$ as a constructor).

@@

First of all, we need a sort of terms which have to be indexed by
types because we only want well-typed terms. Both types and terms can
include variables, and to keep track of the currently available
variables, we also index them by contexts: a context is a list of
types, the length is the number of available variables, and we add new
variables at the end (that is, they are snoc-lists rather than
cons-lists). Types are also indexed by their universe level, this is a
technical requirement for avoiding Russell's paradox. The index $i$ is
an implicit argument of $\Tm$.
\[
\Con : \Set \hspace{3em} 
\Ty : \Con\ra\N\ra\Set \hspace{3em} 
\Tm : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Set
\]
Just as lists have two constructors, there are two ways to form
contexts: the empty context $\diamond$ and context extension $\ext$
which is like the snoc operation for lists. Context extension takes a
type which can have variables in the context preceding the type.
\[
\diamond : \Con \hspace{3em}
\blank\ext\blank : (\Gamma:\Con)\ra\Ty\,\Gamma\,i\ra\Con
\]
We will refer to variables by their distance from the end of the
context. We would like to describe the operation providing the last
variable in a context, the zero De Bruijn index \cite{debruijn}, we
denote it by $\q : \Tm\,(\Gamma\ext A)\,A'$. It is a term in an
extended context, and it should have type $A$, but the issue is that
$A : \Ty\,\Gamma$, and $A' : \Ty\,(\Gamma\ext A)$. We need a way to
weaken the type $A$ to obtain such an $A'$. For this, we will
introduce an operation $\blank[\p] : \Ty\,\Gamma\,i\ra\Ty\,(\Gamma\ext
A)\,i$ and define $A' := A[\p]$. Instead of introducing just
$\blank[\p]$, we generalise a bit and add a new sort $\Sub$ which for
now only has the single element $\p$ and the operation
$\blank[\blank]$ is called \emph{instantiation}. We instantiate the
type with the $\Sub\,\Delta\,\Gamma$ weakening.  Elements of $\Sub$
will later be called substitutions, hence the name, but for now, we
only have single end-of context weakenings in $\Sub$. The context
$\Delta$ is called the domain, $\Gamma$ the codomain of a
$\Sub\,\Delta\,\Gamma$.
\begin{alignat*}{10}
  & \Sub && : \Con\ra\Con\ra\Set && \p && : \Sub\,(\Gamma\ext A)\,\Gamma \\
  & \blank[\blank] && : \Ty\,\Gamma\,i\ra\Sub\,\Delta\,\Gamma\ra\Ty\,\Delta\,i \hspace{3em} && \q && : \Tm\,(\Gamma\ext A)\,(A[\p])
\end{alignat*}
The operations $\p$ and $\q$ take three implicit parameters, $\Gamma$,
$i$ and $A$, while $\blank[\blank]$ takes $\Gamma$, $i$ and $\Delta$
implicitly. Still, we only have the last variable $\q$ in the context,
we don't have e.g.\ the last but one variable in $\Tm\,(\Gamma\ext
A\ext B)\,(A[\p][\p])$ where we had to weaken $A$ twice to make it fit
with its context. To obtain more variables, we also allow weakening of
terms, more precisely, we introduce an instantiation operation for
terms with an overloaded notation.
\begin{alignat*}{10}
& \blank[\blank] && : \Tm\,\Gamma\,A\ra(\gamma:\Sub\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\gamma])
\end{alignat*}
Note that this is a dependent function as the type of the resulting
term has to be weakened the same way as the term itself. Now we can
define all variables counting from the end of the context (De Bruijn
indices): $0 := \q$, $1 := \q[\p]$, $2 := \q[\p][\p]$,
$3 := \q[\p][\p][\p]$, and so on.

@@

We add the rules for universes. A universe is a type containing codes
for types, this is witnessed by $\El$ (elements) and $\cd$ (code),
which make an isomorphism between terms of type $\U\,i$ and types of
level $i$. All three operations come with substitution rules.
% ($\El[]$
% and $\cd[]$ are interderivable, so it would be enough to assume one of
% the two, but we add both to completely align with the algorithm
% \cite{DBLP:conf/fscd/KaposiX24} generating the rules from the SOGAT
% (\ref{eq:tt})).
\begin{alignat*}{10}
  & \U && : (i:\N)\ra\Ty\,\Gamma\,(1+i)\hspace{3em} && \U[] && : (\U\,i)[\gamma] = \U\,i                           && \U\beta && : \El\,(\cd\,A) = A \\        
  & \El && : \Tm\,\Gamma\,(\U\,i) \ra \Ty\,\Gamma\,i && \El[] && : (\El\,\hat{A})[\gamma] = \El\,(\hat{A}[\gamma]) \hspace{3em} && \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A} \\
  & \cd && : \Ty\,\Gamma\,i\ra\Tm\,\Gamma\,(\U\,i) && \cd[] && : (\cd\,A)[\gamma] = \cd\,(A[\gamma])
\end{alignat*}
Finally, we add the rules for moving types one level up. Because $\Pi$
needs two types at the same level, without lifting, we cannot even
define the polymorphic identity function.
\begin{alignat*}{10}
  & \Lift && : \Ty\,\Gamma\,i\ra\Ty\,\Gamma\,(1+i)\hspace{3em} && (\Lift\,A)[\gamma] = \Lift\,(A[\gamma])\hspace{3em} && \Lift\beta && : \un\,(\mk\,a) = a \\
  & \mk && : \Tm\,\Gamma\,A \ra \Tm\,\Gamma\,(\Lift\,A) && (\mk\,a)[\gamma] = \mk\,(a[\gamma])              && \Lift\eta && : \mk\,(\un\,a) = a \\
  & \un && : \Tm\,\Gamma\,(\Lift\,A) \ra \Tm\,\Gamma\,A && (\un\,a)[\gamma] = \un\,(a[\gamma])
\end{alignat*}
Note that we don't have definitional commutation of type formers and
$\Lift$ such as $\Lift\,(A\Ra B) = (\Lift\,A\Ra\Lift\,B)$. It however
holds as a definitional isomorphism (see Subsection
\ref{sec:examples}). This adds extra administration when using our
theory, but makes stating our theory simpler, and includes more
models. The situation is similar to the correspondence between
$\Ty\,\Gamma\,i$ and $\Tm\,\Gamma\,(\U\,i)$ which could have been
stated as an equality (making universes {\`a} la Russell).

@@

The operations for $\U$ and $\Lift$ can be written more concisely as
the following isomorphisms:
\[
\El:\Tm\,\Gamma\,(\U\,i) \cong \Ty\,\Gamma\,i:\cd \hspace{5em} \un:\Tm\,\Gamma\,(\Lift\,A)\cong\Tm\,\Gamma\,A:\mk
\]
The $\U\beta$, $\U\eta$ and $\Lift\beta$, $\Lift\eta$ rules express
that the round-trips are identities. These isomorphisms are also
\emph{natural}, or compatible with substitutions: these are expressed
by the equations $\El[]$, $\un[]$, $\cd[]$, $\mk[]$.

$\Sigma$ types were introduced by a similar isomorphism, and we give
names to the round-trips equations according to our convention introduced in
Section \ref{sec:metatheory}: $\Sigma\beta_1 : \fst\,(a,b) = a$,
$\Sigma\beta_2 : \snd\,(a,b) = b$ and $\Sigma\eta : w =
(\fst\,w,\snd\,w)$. Naturality for $\Sigma$ was only stated in one
direction (${,}[]$), naturalities in the other direction are
derivable, for $\fst$ this is
$
(\fst\,w)[\gamma]  \overset{\Sigma\beta_1}{=}                          
\fst\,\big((\fst\,w)[\gamma],(\snd\,w)[\gamma]\big) \overset{{,}[]}{=} 
\fst\,\big((\fst\,w,\snd\,w)[\gamma]\big)  \overset{\Sigma\eta}=        
\fst\,(w[\gamma]).
$
For $\Pi$ types, we construct a similar isomorphism
$
(\lambda t.t[\p]\cdot\q) : \Tm\,\Gamma\,(\Pi\,A\,B) \cong \Tm\,(\Gamma\ext A)\,B : \lam.
$
The first round-trip equality is proven as
$\beta : (\lam\,t)[\p]\cdot\q \overset{\lam[]}{=} \lam\,(t[\p^+])\cdot\q \overset{\Pi\beta}{=} t[p^+][\langle\q\rangle] \overset{\text{(\ref{eq:lifted4})}}{=} t$.
The last step is the admissible equation (\ref{eq:lifted4}) from
Section \ref{sec:admissible}. As above with weakening into an
arbitrary context, this equation holds in the syntax, but not in an
arbitrary model.
%and also if $t$ is not a metavariable, but an operation, e.g. if $t = \q$ or $t = \q[\p]\cdot\q$).
The other round-trip is exactly $\Pi\eta : \lam\,(t[\p]\cdot\q) = t$.

We construct another isomorphism between terms depending on a variable and
terms depending on the lifted version of the same variable:
\begin{equation}\label{eq:liftvar}
(\lambda t.t[\p^+][\langle\un\,\q\rangle]) : \Tm\,(\Gamma\ext A)\,B \cong \Tm\,(\Gamma\ext\Lift\,A)\,(B[\p^+][\langle\q\rangle]) : (\lambda t.t[\p^+][\langle\mk\,\q\rangle])
\end{equation}
To derive the round-trips, we again need admissible equations. We prove
the first round-trip equality in Appendix \ref{app:liftvar}, the other
is analogous.
Once we have all the above isomorphisms, we can compose them to obtain commutation of $\Lift$ and $\Pi$:
\begin{alignat*}{10}
  & \Tm\,\Gamma\,\big(\Lift\,(\Pi\,A\,B)\big) \overset{\text{$\Lift$}}{\cong}
  \Tm\,\Gamma\,(\Pi\,A\,B) \overset{\text{$\Pi$}}{\cong} 
  \Tm\,(\Gamma\ext A)\,B \overset{\text{$\Lift$}}{\cong} 
  \Tm\,(\Gamma\ext A)\,(\Lift\,B) \overset{\text{lifted var}}{\cong} \\
  & \Tm\,(\Gamma\ext \Lift\,A)\,(\Lift\,B[\p^+][\langle\un\,\q\rangle]) \overset{\text{$\Pi$}}{\cong}
  \Tm\,\Gamma\,\big(\Pi\,(\Lift\,A)\,(\Lift\,B[\p^+][\langle\un\,\q\rangle])\big)
\end{alignat*}
We hope that the above examples demonstrated the need for equations
(\ref{eq:lifted1})--(\ref{eq:lifted4}) which we will prove in the next
section.

@@

We can simulate parallel substitutions using $\Sub^*$ as iterated single
substitutions, however it does not satisfy the equations of parallel
substitutions. Thus we define parallel substitutions as lists of terms. It seems to be
difficult to define the instantiation operation for $\Tms$ directly, so we also define
a map into $\Sub^*$, and reuse the instantiation operation of $\Sub^*$ to
convert it into a sequence of instantiations of single substitutions.
\begin{alignat*}{10}
  & \Tms && : \Con\ra\Con\ra\Con && \llcorner\blank\lrcorner && : \Tms\,\Delta\,\Gamma \to \Sub^*\,\Delta\,\Gamma \\
  & \epsilon && : \Tms\,\Gamma\,\diamond && \llcorner\epsilon\lrcorner && := \id\circ\p\circ\dots\circ\p \\
  & \blank,\blank && : (\gamma:\Tms\,\Delta\,\Gamma)\ra\Tm\,\Delta\,(A[\llcorner\gamma\lrcorner])\ra\Tms\,\Delta\,(\Gamma\ext A)\hspace{1.9em} && \llcorner\gamma,a\lrcorner && := \llcorner\gamma\lrcorner^+\circ\langle a \rangle
\end{alignat*}
The computation of instantiating with $\Tms$ is illustrated below. On the
right-hand side, we first weaken $B$ to the context in which the $a_i$'s live,
then replace the variables of $B$ one by one using the $a_i$'s, which are
well-typed because of the previous weakening.
\[
  B[\llcorner\epsilon,a_1,a_2,\dots,a_{n-1},a_n\lrcorner] = 
  B[\p^{+^n}]\dots[\p^{+^n}][\langle a_1 \rangle^{+^{n-1}}][\langle a_2 \rangle^{+^{n-2}}]\dots[\langle a_{n-1} \rangle^+][\langle a_n \rangle]
\]
All CwF operations and equations can be defined with $\Tms$ by induction, using
Lemma \ref{thm:lifting} to avoid further induction on the syntax.

\begin{problem}
  Contexts, types, and terms in the SSC syntax are isomorphic to the
  corresponding sorts in the CwF syntax. In addition $\Tms$ is isomorphic to CwF
  substitutions.
\end{problem}
\begin{proof}[Construction]\leavevmode
  \begin{description}
    \item[$\Rightarrow$] By recursion on the syntax, SSC operations can be
    trivially interpreted by CwF operations.
    \item[$\Leftarrow$] By recursion on the syntax, using $\Tms$ to interpret
    parallel substitutions.
  \end{description}
  The round-trips are proven by induction on the syntax.
\end{proof}
This also implies that the initial SSC is the initial CwF.

@@

The other round-trip, that is, starting with an SSC-model, termifying

@@

isomorphic after the round-trip. We leave formulating the right notion
